Ecosystem functional response across precipitation extremes in a sagebrush steppe

### 2017-09-25 manuscript/usses_water_ms.Rmd
trivia
final edits before PeerJ submission

  **Results.** Ecosystem functional response, defined as the relationship between soil moisture and ANPP was similar among drought and control treatments, but the irrigation treatment had a lower slope than the control treatment. However, ANPP response to availabile soil moisture was weak and uncertain regardless of treatment, with all slopes overlapping zero. There was also large spatial variation in ANPP within-years. Plant community composition was remarkably stable over the course of the experiment and did not differ among treatments.

@@

At any given site, the relationship between aboveground net primary productivity (ANPP) and water availability (e.g., soil moisture) can be characterized by regressing historical observations of ANPP on observations of soil moisture. 
The fitted functional response can then be used to project ANPP under future precipitation regimes (REFERENCE). 
A problem with this approach is that it requires extrapolation if future precipitation falls outside outside the historical range of variability [@Smith2011; @Peters2012].
For example, the soil moisture-ANPP relationship may be linear within the historical range of interannual variation, but could saturate at higher levels of soil moisture. 
In fact, saturating relationships are actually common [@Hsu2012; @Gherardi2015a], perhaps because other resources, like nitrogen, become more limiting in wet years than dry years.
Failure to accurately estimate the curvature of the soil moisture-ANPP relationship will lead to over- or underprediction of ANNP under extreme precipitation [@Peters2012].

@@

## Study Area
We conducted our precipitation manipulation experiment in a sagebrush steppe community at the USDA, U.S. Sheep Experiment Station (USSES) near Dubois, Idaho (44.2$^{\circ}$ N, 112.1$^{\circ}$ W), 1500 m above sea level.
The plant community is dominated by the shrub *Artemesia tripartita* and three perennial bunchgrasses, *Pseudoroegneria spicata*, *Poa secunda*, and *Hesperostipa comata*.
During the period of our experiment (2011 -- 2015), average mean annual precipitation was 265 mm year$\phantom{}^{-1}$ and mean monthly temperature ranged from -5.2$^{\circ}$C in January to 21.8$^{\circ}$C in July.
Between 1926 and 1932, range scientists at the USSES established 26 permanent 1 m$^2$ quadrats to track vegetation change over time.
In 2007, we relocated 14 of the original quadrats, six of which were inside a large, permanent livestock exclosure.
We use these six plots as control plots (i.e. ambient precipitation) in the experiment described below.

@@

For each plot and year, we averaged the four readings for each wavelength and then calculated a greenness index based on the same bands used to calculate NDVI using the MODIS and AVHRR [YOU HAVEN'T DEFINED THESE] bands for NDVI.
We regressed the greenness index against the dry biomass weight from the ten calibration plots to convert the greenness index to ANPP.
We fit regressions to a MODIS-based index and an AVHRR-based index for each year and retained the regression with the better fit based on $R^2$ values.
We then predicted ANPP using the best regression equation for each year (Appendix 1).

@@

These hypotheses are based on evidence that precipitation-ANPP relationships often saturate with increasing precipitation (REFERENCES).
<!---these should probably be stated in the Intro--->

@@

\noindent{}where $\Sigma$ is the variance-covariance matrix and *j(i)* reads as "plot *j* associated with observation *i*". 
The random year effects ($\boldsymbol{\eta}$) are drawn from a normal prior with mean 0 and standard deviation $\sigma_{\text{year}}$, which was drawn from a half-Cauchy distribution. 
A full description of our model is in Appendix 2.
<!---call out available computer code here?--->

@@

Three of our five study years received below average precipitation (Fig. 2A).
Because our treatments are proportional changes in ambient precipitation, those three years represent a lower magnitude of absolute difference in precipitation among the treatments.
<!---why is the previous point important here? should it go in Methods?--->
ANPP varied from a minimum of 74.5 g m$^{-2}$ in 2014 to a maximum of 237.1 g m$^{-2}$ in 2016 when averaged across treatments (Fig. 2C).
ANPP was slightly higher in irrigation plots and slightly lower in drought plots (Fig. 2C), corresponding to estimated soil volumetric water content (VWC) differences among treatments (Fig. 2B).
Such differences in soil VWC indicate our treatment infrastructure was successful.
ANPP was highly variable across plots within years (Fig. 2C).

@@

Community composition was similar among treatments.
Community composition among treatments overlapped in all years and was equally dispersed in all years (Table 2; Fig. 5).
Community composition was also remarkably stable over time, with no evidence of divergence among treatments (Table 2; Fig. 5).

@@

The results of our five year experiment in a sagebrush steppe conform to two of our four predictions, depending on treatment.
Ecosystem functional response and community composition did not change under chronic drought (Fig. 3A, Fig. 4), representing the top left scenario in Fig. 1.
Ecosystem functional response under chronic irrigation was different from the control treatment, but community composition remained unchanged (Fig. 3A, Fig. 4), representing the top right scenario in Fig. 1.
The altered ecosystem functional response under irrigation matched our expectations because we found evidence for a weaker response for the irrigation treatment (Fig. 3).
Although the difference between the control and irrigation slopes is likely different from zero (Fig. 3A), the posterior distribution of the slopes (not their difference) are similar and overlap broadly (Table1, Fig. A2-2).
Thus, the statistical difference between the slopes is not ecologically meaningful.
We therefore conclude that ecosystem functional response is consistent and weak across all precipitation treatments.

@@

Third, the life history traits of the dominant species in our study ecosystem may explain the consistently positive, but weak and uncertain, effect of soil moisture on ANPP (Table 1, Fig. 3).
Species that live in variable environments, such as cold deserts, must have strategies to ensure long-term success as conditions vary.
One strategy is bet hedging, where species forego short-term gains to reduce the variance of long-term success [@Seger1987].
In other words, species follow the same conservative strategy every year, designed to minimize response to environmental conditions.
The dry and variable environment of the sagebrush steppe has likely selected for bet hedging species that can maintain function at low water availability and have weak responses to high water availability.
In so doing, the dominant species in our ecosystem avoid "boom and bust" cycles, which corresponds to the weak effect of soil moisture on ANPP (i.e., the Bayesian credible intervals for the slopes in $\boldsymbol{\beta}$ in Eq. 1 overlapping zero).

@@

\begin{figure}[!ht]
  \centering
      \includegraphics[height=7in]{../figures/data_panels.png}
  \caption{(A) Probability density of historical precipitation from 1926-2016, with the years of the experiment shown with arrows on the \emph{x}-axis. (B) Mean soil volumetric water content (VWC) in each of the three treatments during the course of the experiment. (C) Mean (points) ANPP and its standard deviation (error bars) for each year of the experiment. **\emph{note plot b doesn't have a treatment legend, consider adding to the side of (B) and removing from (C)} }
\end{figure}

@@

\begin{figure}[!ht]
  \centering
      \includegraphics[width=5in]{../figures/sppcomp_bray_all.png}
  \caption{Nonmetric multidimensional scaling scores representing plant communities in each plot, colored by treatment. **\emph{Worth adding arrow vectors showing association between axes and dominant species?}}
\end{figure}

>>>>

  **Results.** Ecosystem functional response, defined as the relationship between soil moisture and ANPP was similar among drought and control treatments, but the irrigation treatment had a lower slope than the control treatment. However, ANPP response to available soil moisture was weak and uncertain regardless of treatment, with all slopes overlapping zero. There was also large spatial variation in ANPP within-years. Plant community composition was remarkably stable over the course of the experiment and did not differ among treatments.

@@

At any given site, the relationship between aboveground net primary productivity (ANPP) and water availability (e.g., soil moisture) can be characterized by regressing historical observations of ANPP on observations of soil moisture. 
Fitted functional responses can then be used to infer how ANPP may change under future precipitation regimes [e.g., @Hsu2012]. 
A problem with this approach is that it requires extrapolation if future precipitation falls outside outside the historical range of variability [@Smith2011; @Peters2012].
For example, the soil moisture-ANPP relationship may be linear within the historical range of interannual variation, but could saturate at higher levels of soil moisture. 
In fact, saturating relationships are actually common [@Hsu2012; @Gherardi2015a], perhaps because other resources, like nitrogen, become more limiting in wet years than dry years.
Failure to accurately estimate the curvature of the soil moisture-ANPP relationship will lead to over- or underprediction of ANNP under extreme precipitation [@Peters2012].

@@

## Study Area

We conducted our precipitation manipulation experiment in a sagebrush steppe community at the USDA, U.S. Sheep Experiment Station (USSES) near Dubois, Idaho (44.2$^{\circ}$ N, 112.1$^{\circ}$ W), 1500 m above sea level.
The plant community is dominated by the shrub *Artemesia tripartita* and three perennial bunchgrasses, *Pseudoroegneria spicata*, *Poa secunda*, and *Hesperostipa comata*.
During the period of our experiment (2011 -- 2015), average mean annual precipitation was 265 mm year$\phantom{}^{-1}$ and mean monthly temperature ranged from -5.2$^{\circ}$C in January to 21.8$^{\circ}$C in July.
Between 1926 and 1932, range scientists at the USSES established 26 permanent 1 m$^2$ quadrats to track vegetation change over time.
In 2007, we relocated 14 of the original quadrats, six of which were inside a large, permanent livestock exclosure.
We use these six plots as control plots (i.e. ambient precipitation) in the experiment described below.

@@

For each plot and year, we averaged the four readings for each wavelength and then calculated a greenness index based on the same bands used to calculate NDVI using the MODIS (Moderate Resolution Imaging Spectroradiometer) and AVHRR (Advanced Very High Resolution Radiometer) bands for NDVI.
We regressed the greenness index against the dry biomass weight from the ten calibration plots to convert the greenness index to ANPP.
We fit regressions to a MODIS-based index and an AVHRR-based index for each year and retained the regression with the better fit based on $R^2$ values.
We then predicted ANPP using the best regression equation for each year (Appendix 1).

@@

\noindent{}These hypotheses are based on evidence that precipitation-ANPP relationships often saturate with increasing precipitation [@Hsu2012; @Gherardi2015a].

@@

\noindent{}where $\Sigma$ is the variance-covariance matrix and *j(i)* reads as "plot *j* associated with observation *i*". 
The random year effects ($\boldsymbol{\eta}$) are drawn from a normal prior with mean 0 and standard deviation $\sigma_{\text{year}}$, which was drawn from a half-Cauchy distribution. 
A full description of our model and associated R [@R2016] code is in Appendix 2.

@@

Ambient precipitation was variable over the five years of the study (Fig. 2A), meaning our experiment captured a range of environmental conditions.
ANPP varied from a minimum of 74.5 g m$^{-2}$ in 2014 to a maximum of 237.1 g m$^{-2}$ in 2016 when averaged across treatments (Fig. 2C).
ANPP was slightly higher in irrigation plots and slightly lower in drought plots (Fig. 2C), corresponding to estimated soil volumetric water content (VWC) differences among treatments (Fig. 2B).
Such differences in soil VWC indicate our treatment infrastructure was successful.
ANPP was highly variable across plots within years (Fig. 2C).

@@

Community composition was similar among treatments.
The multidimensional space of community composition overlapped among treatments in all years and was equally dispersed in all years (Table 2; Fig. 5).
Community composition was also remarkably stable over time, with no evidence of divergence among treatments (Table 2; Fig. 5).

@@

The results of our five year experiment in a sagebrush steppe conform to two of our four predictions, depending on treatment.
Ecosystem functional response and community composition did not change under chronic drought (Fig. 3A, Fig. 4), representing the top left scenario in Fig. 1.
Ecosystem functional response under chronic irrigation was different from the control treatment, but community composition remained unchanged (Fig. 3A, Fig. 4), representing the top right scenario in Fig. 1.
The decrease in the slope of the VWC $\times$ productivity relationship in the irrigated plots is consistent with a weaker response of productivity to additional water in this ecosystem.
This suggests that if average soil moisture were pushed consistently higher than currently observed ambient conditions, there would be a weaker relationship between precipitation and productivity in this system.
Nonetheless, the slopes of the VWC $\times$ productivity relationship were similar among treatments (Table 1), at least at the level of irrigation we imposed.
We therefore conclude that ecosystem functional response is consistent and weak across all precipitation treatments.

@@

Third, the life history traits of the dominant species in our study ecosystem may explain the consistently positive, but weak and uncertain, effect of soil moisture on ANPP (Table 1, Fig. 3).
Species that live in variable environments, such as cold deserts, must have strategies to ensure long-term success as conditions vary.
One strategy is bet hedging, where species forego short-term gains to reduce the variance of long-term success [@Seger1987].
In other words, species follow the same conservative strategy every year, designed to minimize response to environmental conditions.
The dry and variable environment of the sagebrush steppe has likely selected for bet hedging species that can maintain function at low water availability and have weak responses to high water availability.
In so doing, the dominant species in our ecosystem avoid "boom and bust" cycles, which corresponds to the weak effect of soil moisture on ANPP (i.e., the Bayesian credible intervals for the slopes overlapping zero; Table 1).

@@

\begin{figure}[!ht]
  \centering
      \includegraphics[height=7in]{../figures/data_panels.png}
  \caption{(A) Probability density of historical precipitation from 1926-2016, with the years of the experiment shown with arrows on the \emph{x}-axis. (B) Mean soil volumetric water content (VWC) in each of the three treatments during the course of the experiment. (C) Mean (points) ANPP and its standard deviation (error bars) for each year of the experiment. Colors in panels B and C identify the treatment, as specified in the legend of panel C.}
\end{figure}

@@

\begin{figure}[!ht]
  \centering
      \includegraphics[width=5in]{../figures/sppcomp_bray_all.png}
  \caption{Nonmetric multidimensional scaling scores representing plant communities in each plot, colored by treatment.}
\end{figure}

### 2017-09-25 manuscript/usses_water_ms.tex
final edits before PeerJ submission

\textbf{Results.} Ecosystem functional response, defined as the
relationship between soil moisture and ANPP was similar among drought
and control treatments, but the irrigation treatment had a lower slope
than the control treatment. However, ANPP response to availabile soil
moisture was weak and uncertain regardless of treatment, with all slopes
overlapping zero. There was also large spatial variation in ANPP
within-years. Plant community composition was remarkably stable over the
course of the experiment and did not differ among treatments.

@@

At any given site, the relationship between aboveground net primary
productivity (ANPP) and water availability (e.g., soil moisture) can be
characterized by regressing historical observations of ANPP on
observations of soil moisture. The fitted functional response can then
be used to project ANPP under future precipitation regimes (REFERENCE).
A problem with this approach is that it requires extrapolation if future
precipitation falls outside outside the historical range of variability
\citep{Smith2011, Peters2012}. For example, the soil moisture-ANPP
relationship may be linear within the historical range of interannual
variation, but could saturate at higher levels of soil moisture. In
fact, saturating relationships are actually common
\citep{Hsu2012, Gherardi2015a}, perhaps because other resources, like
nitrogen, become more limiting in wet years than dry years. Failure to
accurately estimate the curvature of the soil moisture-ANPP relationship
will lead to over- or underprediction of ANNP under extreme
precipitation \citep{Peters2012}.

@@

For each plot and year, we averaged the four readings for each
wavelength and then calculated a greenness index based on the same bands
used to calculate NDVI using the MODIS and AVHRR {[}YOU HAVEN'T DEFINED
THESE{]} bands for NDVI. We regressed the greenness index against the
dry biomass weight from the ten calibration plots to convert the
greenness index to ANPP. We fit regressions to a MODIS-based index and
an AVHRR-based index for each year and retained the regression with the
better fit based on \(R^2\) values. We then predicted ANPP using the
best regression equation for each year (Appendix 1).

@@

These hypotheses are based on evidence that precipitation-ANPP
relationships often saturate with increasing precipitation (REFERENCES).

@@

\noindent{}where \(\Sigma\) is the variance-covariance matrix and
\emph{j(i)} reads as ``plot \emph{j} associated with observation
\emph{i}''. The random year effects (\(\boldsymbol{\eta}\)) are drawn
from a normal prior with mean 0 and standard deviation
\(\sigma_{\text{year}}\), which was drawn from a half-Cauchy
distribution. A full description of our model is in Appendix 2.

@@

Three of our five study years received below average precipitation (Fig.
2A). Because our treatments are proportional changes in ambient
precipitation, those three years represent a lower magnitude of absolute
difference in precipitation among the treatments. ANPP varied from a
minimum of 74.5 g m\(^{-2}\) in 2014 to a maximum of 237.1 g m\(^{-2}\)
in 2016 when averaged across treatments (Fig. 2C). ANPP was slightly
higher in irrigation plots and slightly lower in drought plots (Fig.
2C), corresponding to estimated soil volumetric water content (VWC)
differences among treatments (Fig. 2B). Such differences in soil VWC
indicate our treatment infrastructure was successful. ANPP was highly
variable across plots within years (Fig. 2C).

@@

The results of our five year experiment in a sagebrush steppe conform to
two of our four predictions, depending on treatment. Ecosystem
functional response and community composition did not change under
chronic drought (Fig. 3A, Fig. 4), representing the top left scenario in
Fig. 1. Ecosystem functional response under chronic irrigation was
different from the control treatment, but community composition remained
unchanged (Fig. 3A, Fig. 4), representing the top right scenario in Fig.
1. The altered ecosystem functional response under irrigation matched
our expectations because we found evidence for a weaker response for the
irrigation treatment (Fig. 3). Although the difference between the
control and irrigation slopes is likely different from zero (Fig. 3A),
the posterior distribution of the slopes (not their difference) are
similar and overlap broadly (Table1, Fig. A2-2). Thus, the statistical
difference between the slopes is not ecologically meaningful. We
therefore conclude that ecosystem functional response is consistent and
weak across all precipitation treatments.

@@

Third, the life history traits of the dominant species in our study
ecosystem may explain the consistently positive, but weak and uncertain,
effect of soil moisture on ANPP (Table 1, Fig. 3). Species that live in
variable environments, such as cold deserts, must have strategies to
ensure long-term success as conditions vary. One strategy is bet
hedging, where species forego short-term gains to reduce the variance of
long-term success \citep{Seger1987}. In other words, species follow the
same conservative strategy every year, designed to minimize response to
environmental conditions. The dry and variable environment of the
sagebrush steppe has likely selected for bet hedging species that can
maintain function at low water availability and have weak responses to
high water availability. In so doing, the dominant species in our
ecosystem avoid ``boom and bust'' cycles, which corresponds to the weak
effect of soil moisture on ANPP (i.e., the Bayesian credible intervals
for the slopes in \(\boldsymbol{\beta}\) in Eq. 1 overlapping zero).

>>>>

\textbf{Results.} Ecosystem functional response, defined as the
relationship between soil moisture and ANPP was similar among drought
and control treatments, but the irrigation treatment had a lower slope
than the control treatment. However, ANPP response to available soil
moisture was weak and uncertain regardless of treatment, with all slopes
overlapping zero. There was also large spatial variation in ANPP
within-years. Plant community composition was remarkably stable over the
course of the experiment and did not differ among treatments.

@@

At any given site, the relationship between aboveground net primary
productivity (ANPP) and water availability (e.g., soil moisture) can be
characterized by regressing historical observations of ANPP on
observations of soil moisture. Fitted functional responses can then be
used to infer how ANPP may change under future precipitation regimes
\citep[e.g.,][]{Hsu2012}. A problem with this approach is that it
requires extrapolation if future precipitation falls outside outside the
historical range of variability \citep{Smith2011, Peters2012}. For
example, the soil moisture-ANPP relationship may be linear within the
historical range of interannual variation, but could saturate at higher
levels of soil moisture. In fact, saturating relationships are actually
common \citep{Hsu2012, Gherardi2015a}, perhaps because other resources,
like nitrogen, become more limiting in wet years than dry years. Failure
to accurately estimate the curvature of the soil moisture-ANPP
relationship will lead to over- or underprediction of ANNP under extreme
precipitation \citep{Peters2012}.

@@

For each plot and year, we averaged the four readings for each
wavelength and then calculated a greenness index based on the same bands
used to calculate NDVI using the MODIS (Moderate Resolution Imaging
Spectroradiometer) and AVHRR (Advanced Very High Resolution Radiometer)
bands for NDVI. We regressed the greenness index against the dry biomass
weight from the ten calibration plots to convert the greenness index to
ANPP. We fit regressions to a MODIS-based index and an AVHRR-based index
for each year and retained the regression with the better fit based on
\(R^2\) values. We then predicted ANPP using the best regression
equation for each year (Appendix 1).

@@

\noindent{}These hypotheses are based on evidence that
precipitation-ANPP relationships often saturate with increasing
precipitation \citep{Hsu2012, Gherardi2015a}.

@@

\noindent{}where \(\Sigma\) is the variance-covariance matrix and
\emph{j(i)} reads as ``plot \emph{j} associated with observation
\emph{i}''. The random year effects (\(\boldsymbol{\eta}\)) are drawn
from a normal prior with mean 0 and standard deviation
\(\sigma_{\text{year}}\), which was drawn from a half-Cauchy
distribution. A full description of our model and associated R
\citep{R2016} code is in Appendix 2.

@@

Ambient precipitation was variable over the five years of the study
(Fig. 2A), meaning our experiment captured a range of environmental
conditions. ANPP varied from a minimum of 74.5 g m\(^{-2}\) in 2014 to a
maximum of 237.1 g m\(^{-2}\) in 2016 when averaged across treatments
(Fig. 2C). ANPP was slightly higher in irrigation plots and slightly
lower in drought plots (Fig. 2C), corresponding to estimated soil
volumetric water content (VWC) differences among treatments (Fig. 2B).
Such differences in soil VWC indicate our treatment infrastructure was
successful. ANPP was highly variable across plots within years (Fig.
2C).

@@

The results of our five year experiment in a sagebrush steppe conform to
two of our four predictions, depending on treatment. Ecosystem
functional response and community composition did not change under
chronic drought (Fig. 3A, Fig. 4), representing the top left scenario in
Fig. 1. Ecosystem functional response under chronic irrigation was
different from the control treatment, but community composition remained
unchanged (Fig. 3A, Fig. 4), representing the top right scenario in Fig.
1. The decrease in the slope of the VWC \(\times\) productivity
relationship in the irrigated plots is consistent with a weaker response
of productivity to additional water in this ecosystem. This suggests
that if average soil moisture were pushed consistently higher than
currently observed ambient conditions, there would be a weaker
relationship between precipitation and productivity in this system.
Nonetheless, the slopes of the VWC \(\times\) productivity relationship
were similar among treatments (Table 1), at least at the level of
irrigation we imposed. We therefore conclude that ecosystem functional
response is consistent and weak across all precipitation treatments.

@@

Third, the life history traits of the dominant species in our study
ecosystem may explain the consistently positive, but weak and uncertain,
effect of soil moisture on ANPP (Table 1, Fig. 3). Species that live in
variable environments, such as cold deserts, must have strategies to
ensure long-term success as conditions vary. One strategy is bet
hedging, where species forego short-term gains to reduce the variance of
long-term success \citep{Seger1987}. In other words, species follow the
same conservative strategy every year, designed to minimize response to
environmental conditions. The dry and variable environment of the
sagebrush steppe has likely selected for bet hedging species that can
maintain function at low water availability and have weak responses to
high water availability. In so doing, the dominant species in our
ecosystem avoid ``boom and bust'' cycles, which corresponds to the weak
effect of soil moisture on ANPP (i.e., the Bayesian credible intervals
for the slopes overlapping zero; Table 1).
